---
title: "Spatial lauches"
author: "Mathilde Mousset"
date: "22 janvier 2019"
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

Readme: *https://github.com/rfordatascience/tidytuesday/blob/master/data/2019/2019-01-15/readme.md*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(countrycode)
library(viridis)
theme_set(theme_light())

# Import data from github

agencies <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-15/agencies.csv")

launches <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-15/launches.csv")


# Get better date format in lauches
launches <- launches %>% 
  mutate(launch_date = ymd(launch_date))

# Improve countries from agencies
launches$agency %>% unique()
levels(launches$state_code)

launches <- launches %>% 
  mutate(state_code_clean = fct_collapse(state_code,
    "RU" = c("SU", "RU"),
    "FR" = c("F", "I-ELDO", "I-ESA"),
    "JP" = "J",
    "IT" = "I",
    "KY" = "CYM",
    "GB" = "UK"
  )) %>%
  mutate(state_name = countrycode(state_code_clean, "iso2c", "country.name"),
         state_name_short = fct_lump(state_name, 6)) %>%
  replace_na(list(state_name_short = "Other")) %>% 
  mutate(state_bloc = fct_collapse(state_code_clean,
                                   "URSS"   = "RU",
                                   "Europe" = c("FR", "IT", "GB", "KY"),
                                   "North_america"     = "US",
                                   "Asia"   = c("CN", "JP", "IN", "KR", "KP"),
                                   "South_america" = "BR",
                                   "Middle_east" = c("IR", "IL")))  %>% 
  mutate(category = fct_recode(category,
                                Success = "O",
                                Failure = "F"))
```

# Lauches per country

```{r}
launches %>% 
  count(state_bloc, sort = TRUE) %>% 
  ggplot(aes(x = state_bloc, y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of lauches per state blocs",
       x = "",
       y = "Number of lauches")
```

The same per country
```{r}
launches %>% 
  count(state_name, state_bloc) %>% 
  ggplot(aes(x = state_name, y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of lauches per state",
       x = "",
       y = "Number of lauches") +
  facet_wrap(~state_bloc, scales = "free")
```


```{r}
launches %>% 
  count(launch_year, state_name, state_bloc) %>% 
  ggplot(aes(x = launch_year, y = n, colour = state_bloc)) +
  geom_line() +
  labs(title = "Number of lauches per state",
       x = "",
       y = "Number of lauches") +
  facet_wrap(~state_bloc, scales = "free") +
  facet_wrap(~state_name)
```


# Launches with time

The United states begin the space race, closely followed by Russia, which dominates the number of lauch until the late nineties. After that, Russia and the US are stable and equivalent. In the early 2000, China, followed by India increase their shares of the launches.


```{r}
launches %>% 
  count(state_name, launch_year) %>% 
  mutate(state_name = fct_reorder(state_name, -n, sum)) %>% 
  ggplot(aes(x = launch_year, y = n, colour = state_name)) +
  geom_line(size = 1.5) +
  labs(title = "Number of lauches per state",
       x = "Year of launch",
       y = "Number of lauches",
       colour = "State") 


launches %>% 
  count(state_name_short, launch_year) %>%
  mutate(state_name_short = fct_reorder(state_name_short, -n, sum)) %>% 
  ggplot(aes(x = launch_year, y = n, 
             colour = state_name_short)) +
  geom_line(size = 1.5) +
  labs(title = "Number of lauches per state (short list)",
       x = "Year of launch",
       y = "Number of lauches",
       colour = "State")

launches %>% 
  count(state_bloc, launch_year) %>%
  mutate(state_bloc = fct_reorder(state_bloc, -n, sum)) %>% 
  ggplot(aes(x = launch_year, y = n, 
             colour = state_bloc)) +
  geom_line(size = 1.5) +
  labs(title = "Number of lauches per state bloc",
       subtitle = "US and USSR dominated the early race, Asia is booming now",
       x = "Year of launch",
       y = "Number of lauches",
       colour = "State bloc") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```


# Private vs public lauches

By bloc

```{r}
launches %>% 
  filter(state_bloc != "South_america" & state_bloc != "Middle_east") %>% 
  count(state_bloc, launch_year, agency_type) %>%
  ggplot(aes(x = launch_year, y = n, 
             colour = agency_type)) +
  geom_line(size = 1.5) +
  facet_wrap(~state_bloc, nrow = 4) +
  labs(title = "Number of lauches per state bloc",
       subtitle = "US and USSR dominated the early race, Asia is booming now",
       x = "Year of launch",
       y = "Number of lauches",
       colour = "State bloc") +
  scale_fill_brewer(palette = "Set1") +
   theme_minimal() +
   theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold"))


launches %>% 
  filter(state_bloc != "South_america" & state_bloc != "Middle_east") %>% 
  count(state_bloc, launch_year, agency_type) %>%
  ggplot(aes(x = launch_year, y = n, 
             fill = agency_type)) +
   geom_bar(stat ="identity", alpha = 0.7) +
  facet_wrap(~state_bloc,  nrow = 4) +
  labs(title = "Number of lauches per state bloc",
       subtitle = "US and USSR dominated the early race, Asia is booming now",
       x = "Year of launch",
       y = "Number of lauches",
       colour = "State bloc") +
  scale_fill_brewer(palette = "Set1") +
   theme_minimal() +
   theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold"))

```


```{r}
launches %>% 
  filter(state_bloc != "South_america" & state_bloc != "Middle_east") %>% 
  select(state_bloc, launch_year, agency_type) %>% 
  group_by(state_bloc, launch_year) %>% 
  add_tally() %>% 
  add_count(agency_type) %>% 
  distinct(state_bloc, launch_year, agency_type, n, nn) %>%  
  mutate(proportion = nn / n) %>% 
  ggplot(aes(x = launch_year, y = proportion, 
             fill = agency_type)) +
  geom_bar(stat = "identity") +
  facet_wrap(~state_bloc, nrow = 4) +
  labs(title = "Number of lauches per state bloc",
       subtitle = "US and USSR dominated the early race, Asia is booming now",
       x = "Year of launch",
       y = "Number of lauches",
       Fill = "State bloc") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
   theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold"))
```


```{r}
launches %>% 
  filter(state_bloc != "South_america" & state_bloc != "Middle_east") %>% 
  #count(state_bloc, agency_type, launch_year) %>% 
  ggplot(aes(x = launch_year, fill = agency_type) ) +
  geom_histogram(bins = 62) + 
  facet_wrap(vars(state_bloc), nrow = 4) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold")) +
    scale_fill_brewer(palette = "Set1") +
  labs(title = "Who accounts for space launches in different blocs?",
       subtitle = "While in Europe and in the states the market has mostly been privatised, state accounts for most lauches in Asia and Russia",
       x = "Year of launch",
       y = "Number of launches")
```


# US vs Russia

```{r}
launches %>% 
  filter(state_name %in% c("United States", "Russia")) %>%
   ggplot(aes(x = launch_year, fill = state_name)) +
  geom_area(stat = "bin", 
            position = "identity",
            bins = 30, alpha = 0.6) +
  scale_fill_manual(values = c("#999999", "#E69F00")) +
    theme_minimal() +
  theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold")) +
  labs(title = "Space race: US vs Russia",
       subtitle = "Numer of launches per year in Russia and the states",
       x = "Year of launches ",
       y = "Number of launches",
       fill = "")
```


# Success vs Failure

We also have the information on whether the launch was a success or a failure. Sweet! 

Success vs failure accross time

```{r}
launches %>% 
  count(category, launch_year) %>% 
  ggplot(aes(x = launch_year, y = n,
             fill = category)) +
  geom_bar(stat = "identity", 
            #position = "identity",
            ) +
  scale_fill_manual(values = c("#999999", "#E69F00")) +
    theme_minimal() +
  theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold")) +
  labs(title = "Number of successfull and failed launches accross time",
       subtitle = "It seems the number of failed launches decrease with time",
       x = "Year of launches ",
       y = "Number of launches",
       fill = "Launch outcome")
```

```{r}
launches %>% 
  filter(state_name %in% c("China", "France", "Russia", "United States")) %>% 
  count(category, launch_year, state_name_short) %>% 
  ggplot(aes(x = launch_year, y = n,
             fill = category)) +
  geom_bar(stat = "identity") +
  facet_wrap(~state_name_short, nrow = 4) +
  scale_fill_manual(values = c("#999999", "#E69F00")) +
    theme_minimal() +
  theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold")) +
  labs(title = "Number of successfull and failed launches accross time",
       subtitle = "It seems the number of failed launches decrease with time",
       x = "Year of launches ",
       y = "Number of launches",
       fill = "Launch outcome")
```

It's nice, but I would have to have a look at the proportion of failed launches.

```{r}
launches %>% 
  filter(state_name %in% c("China", "France", "Russia", "United States")) %>% 
  group_by(state_name, launch_year) %>% 
  add_tally() %>% 
  add_count(category) %>% 
  distinct(state_bloc, launch_year, category, n, nn) %>%  
  mutate(proportion = nn / n) %>% 
  ggplot(aes(x = launch_year, y = proportion*100, 
             fill = category)) +
  geom_bar(stat = "identity") +
  facet_wrap(~state_name, nrow = 4) +
  labs(title = "Space mastery increases with time",
       subtitle = "Countries experienced many failures in their early years. Nowadays, lauches are mostly successfull",
       x = "Year of launch",
       y = "Percentage of lauches",
       fill = "Launch outcome") +
  scale_fill_manual(values = c("#999999", "#E69F00")) +
    theme_minimal() +
  theme_minimal() +
   theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold")) +
  scale_x_continuous("", breaks = seq(1960, 2020, 10))
```

