---
title: "TidyTuesday37"
author: "Tam"
date: "22 january 2019"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)  # For data wrangling and visualization
library(janitor)    # For data cleaning
library(lubridate)  # deal with dates
library(viridis)

set.seed(122555)     # For reproducibility
```


I download the data from *https://data.cityofnewyork.us/api/views/43nn-pn8j/rows.csv*.
```{r clean data}
# Get the raw data, clean them a bit

# You can use this url to download the data directly into R (will take a few seconds)
#df <- read_csv("https://data.cityofnewyork.us/api/views/43nn-pn8j/rows.csv")

df <- read_delim("../tidy_thuesday_37/DOHMH_New_York_City_Restaurant_Inspection_Results.csv",
                 delim = ";",
                 na = c("", "NA", "N/A"))


# Some cleaning

df <- df %>% 
  # Clean names
  janitor::clean_names() %>%
  # Get rid of variables I won't use
  select(-phone, -grade_date, -record_date, -building, -street, zipcode) %>% 
  # Rename some variables
  rename(restaurant_id = camis,
         borough = boro) %>% 
  # Replace "Missing" by NA in borough
  mutate(borough = case_when(
    borough == "Missing" ~ NA_character_,
    TRUE ~ borough
  )) %>% 
  # Filter out the restaurants that have not been inspected yet
  filter(inspection_date != "01/01/1900") %>% 
  # Massages the inspection dates
  mutate(inspection_date = mdy(inspection_date),
         inspection_year = year(inspection_date),
         inspection_month = month(inspection_date),
         inspection_floored = floor_date(inspection_date, unit = "month"))  %>% 
  # Add a column saying wether the inspection was gradable
  mutate(gradable = case_when(
    inspection_type %in% c("Cycle Inspection / Initial Inspection", "Cycle Inspection / Re-Inspection", "Pre-Permit (Operational) / Initial Inspection", "Pre-Permit (Operational) / Re-Inspection") & 
      action %in% c("Violations were cited in the following area(s).","No violations were recorded at the time of this inspection.","Establishment Closed by DOHMH") & 
      inspection_date >="27/07/2010" ~ "yes",
  TRUE ~ "no")) %>% 
  mutate(presence_vermin = str_detect(violation_description, "mice|rats|roaches|flies"))


# Save csv clean version
write_csv(df, "../tidy_thuesday_37/nyc_restaurants.csv")
```


# Cuisine description

There are a lot of cuisine descriptions. Let's create a simplified version or two. I will try to aggreagte types of cooking per continent. Roughly...

By continent  

```{r}
unique(df$cuisine_description) %>%  head(10)

df <- df %>% 
  mutate(cuisine_continent = case_when(
    cuisine_description %in% c("Irish", "Pizza", "Pizza/Italian", 
                               "Eastern European", "Spanish",  "French", 
                               "Italian", "Mediterranean", "Tapas", "Greek",
                               "Polish" , "Continental", "German",
                               "Pancakes/Waffles" , "Czech", "English",
                               "Portuguese", "Scandinavian",  
                               "Basque", "Russian") ~ "european",
    
    cuisine_description %in% c("American", "Donuts", "Hamburgers",
                               "Bagels/Pretzels" , "Steak", "Cajun",
                               "Californian", "Hotdogs/Pretzels", 
                               "Barbecue") ~ "north_american",
    
    cuisine_description %in% c("Chinese", "Asian", "Chinese/Japanese","Thai",
                               "Korean", "Indian", "Japanese",
                               "Vietnamese/Cambodian/Malaysia",
                               "Southwestern") ~ "eastern_asian",
   
     cuisine_description %in% c("Middle Eastern",  "Turkish",
                                "Pakistani", "Bangladeshi", "Delicatessen",
                                "Armenian" ,"Afghan", "Moroccan", "Egyptian",
                                "Iranian") ~ "middle_east",
    
    cuisine_description %in% c("Creole", "African", "Creole/Cajun", 
                               "Ethiopian") ~ "african",
    
    cuisine_description %in% c("Mexican", "Caribbean", 
                               "Latin (Cuban, Dominican, Puerto Rican, South & Central American)" , 
                               "Tex-Mex", "Brazilian", "Peruvian", "Chilean") ~ "latin_american",
    
    cuisine_description %in% c("Bakery", "Café/Coffee/Tea", "Soul Food",
                               "Jewish/Kosher" , "Chicken" , "Sandwiches", 
                               "Sandwiches/Salads/Mixed Buffet", "Ice Cream, Gelato, Yogurt, Ices",  "Juice, Smoothies, Fruit Salads" , "Salads",
                               "Australian", "Chinese/Cuban","Soups",
                               "Bottled beverages, including water, sodas, juices, etc.", 
                               "Vegetarian", "Soups & Sandwiches",  
                               "Not Listed/Not Applicable" , "Hotdogs",
                               "Fruits/Vegetables" ,"Nuts/Confectionary") ~ "other",
  
    cuisine_description %in% c("Polynesian", "Hawaiian",  "Seafood", "Indonesian", "Filipino") ~ "polynesian"
  ))
```



# Exploration grades

```{r}
df %>% 
  mutate(grade = fct_collapse(grade, N = "Not Yet Graded")) %>% 
  ggplot(aes(grade, fill = critical_flag)) +
  geom_bar(position = "fill") +
  facet_wrap(~ cuisine_continent) +
    labs(x = "Grade",
         y = "Percentage",
         title = "NYC Restaurant Inspections: 2010 to 2014",
         fill = "Status") +
  theme_light() +
  scale_fill_viridis(option = "viridis", discrete = TRUE)


df %>% 
  filter(grade %in% c("A", "B", "C")) %>% 
  ggplot(aes(grade, fill = critical_flag)) +
  geom_bar(position = "fill") +
    labs(x = "Grade",
         y = "Percentage",
         title = "NYC Restaurant Inspections: 2010 to 2014",
         fill = "Status") +
  theme_light() +
  scale_fill_viridis(option = "viridis", discrete = TRUE)
```


```{r}
df %>%
  select(score, grade) %>%
  drop_na(score, grade) %>%
  group_by(grade) %>%
  filter(grade %in% c("A", "B", "C")) %>% 
  summarize(count  = n(),
            min    = min(score),
            mean   = mean(score),
            median = median(score),
            max    = max(score))
```




# Number of inspection per restaurant

Some restaurants have been 
```{r}
df %>% 
  group_by(restaurant_id) %>% 
  count(restaurant_id, sort = TRUE) %>% 
  head(20)


df %>% 
  group_by(restaurant_id) %>% 
  count(restaurant_id, sort = TRUE) %>% 
  ggplot(aes(x = n)) +
  geom_histogram() +
  labs(x = "Number of inspections for each restaurant",
       title = "Number of inspections per restaurant") +
  theme_light() +
  scale_fill_viridis(option = "viridis", discrete = TRUE)
```

Does it depends on the borough?

```{r}
df %>% 
  group_by(restaurant_id, borough) %>% 
  drop_na(borough) %>% 
  count(restaurant_id, sort = TRUE) %>% 
  ggplot(aes(x = n)) +
  geom_histogram() +
  facet_wrap(~borough) +
  labs(x = "Number of inspections for each restaurant and borough",
       title = "Number of inspections per restaurant") +
  theme_light()
```

And per type of cooking ?

```{r}
df %>% 
  group_by(restaurant_id, cuisine_continent) %>% 
  drop_na(cuisine_continent) %>% 
  count(restaurant_id, sort = TRUE) %>% 
  ggplot(aes(x = n)) +
  geom_histogram() +
  facet_wrap(~cuisine_continent, scales = "free_y") +
  labs(x = "Number of inspections for each restaurant",
       title = "Number of inspections per restaurant") +
  theme_light()
```

There are restaurants that were visited *a lot* for all types f cooking and borough.


# Most recent inspection

Let's have a look at the most recent inspection for each restaurant.

```{r}
# Select gradable ispection, and then select the last inspection
most_recent_inspection <- df %>% 
  filter(gradable == "yes") %>% 
  group_by(restaurant_id) %>% 
  slice(which.max(inspection_date)) %>% 
  ungroup()

# Check
most_recent_inspection %>% 
  group_by(restaurant_id) %>% 
  count(sort = TRUE) # ok
```

On the most recent inspection, the grade is either A, not yet graded, Z or NA. This seem to be normal given how the grading system works.

The A grade have scores lower than 14. I am surprised by the fact that the NA grade have scores that would yield an A grade. Maybe I should go and read how the system works, it could be normal (like, pending grade attibution, or another visit).

```{r}
most_recent_inspection %>%
  drop_na(score) %>% 
  ggplot(aes(x = score)) +
  geom_histogram() +
  facet_wrap(~grade, scales = "free") +
  labs(title = "Scores distribution for each grade",
       subtitle = "Most recent inspection") +
  theme_light()
```

Let's have a look at the critical flag categories. It's a bit fuzzy what it means, they say that critial means "moe likely to contribute to foodborne illness".

```{r}
most_recent_inspection %>% 
  filter(critical_flag != "Not Applicable") %>% 
  ggplot(aes(x = score, fill = critical_flag)) +
  geom_histogram() +
  facet_wrap(~critical_flag) +
  labs(title = "Scores distribution for critical and not critical flags",
       subtitle = "Critical = more likely to contribute to foodborne illness",
       fill = "Critical flag",
       x = "Restaurant score") +
  theme_light() +
  scale_fill_manual(values = c("darkred", "darkgreen"))
```

I am surprised that the distribution of scores look so alike in the two categories... It is true that one can get points (i.e. points linked to violation) with things not related to vermin, but I am a bit surprised.

# Per borough

Are there borough with restaurant with higher scores (or grades) ?

Distribution of score per borough.

```{r}
most_recent_inspection %>% 
  ggplot(aes(x = score)) +
  geom_histogram() +
  facet_wrap(~borough) +
  labs(title = "Scores distribution for each borough",
       subtitle = "Most recent inspection") +
  theme_light()
```

Seems the distribution is bimodal for all boroughs. So the mean is not going to be a decent summary. I will look at the proportion of grades in each borough instead.

Going back on the "all gradable inspections" (not only the recent ones), I get the proportion of each grade for each borough.

It seems that the distribution of grades in each borrogh is about the same (had to check the numbers to convince myself that it was not a mistake). About 80% of the inspection have a "A" grade.

```{r}
df %>% 
  select(borough, grade) %>% 
  drop_na(borough, grade) %>% 
  group_by(borough) %>% 
  add_tally() %>% 
  add_count(borough, grade) %>% 
  distinct(borough, grade, n, nn) %>%
  mutate(prop = nn / n) %>% 
  ggplot(aes(x = grade, y = prop, colour = grade)) +
  geom_point(size = 2) +
  facet_wrap(~borough) +
  labs(title = "Proportion of each grade in each borough") +
  theme_light() +
  scale_colour_viridis(option = "viridis", discrete = TRUE)
```

So much about the idea that some neighbourhoods have restaurants with more problems.


# Violation

I am interested in critical violations. There are about 100 types of violation.

```{r}
df$violation_description %>% unique() %>% head(20)
```

```{r}
df %>% 
  filter(critical_flag == "Critical") %>% 
  select(violation_description) %>% distinct()
```

Let's have a look at restaurants with vermin problems
```{r}
df %>% 
  filter(presence_vermin == TRUE) %>% 
  ggplot(aes(x = score, fill = grade)) +
  facet_wrap(~grade, scales = "free") +
  geom_histogram() +
  scale_fill_viridis(option = "viridis", discrete = TRUE) +
  labs(title = "Distribution of scores in restaurants with vermin",
       subtitle = "Vermin = rats, mice, croaches or flies")
```


```{r}
df %>% 
  select(grade, presence_vermin, score) %>% 
  filter(presence_vermin == TRUE) %>% 
  add_tally() %>% 
  add_count(grade) %>% 
  mutate(prop = nn/n) %>% 
  distinct(grade, prop) %>% 
  ggplot(aes(x = "", y = prop, fill = grade)) +
    geom_bar(width = 1, stat = "identity") +
  labs(title = "Proportion of each grade in restaurants with vermin",
       y = "proportion",
       subtitle = "Restaurants with vermin related violations") +
  theme_light() +
  scale_fill_viridis(option = "viridis", discrete = TRUE)
```

There is a quite decent proportion of restaurants that get an A despite the presence of mouse, flies, croaches or rats.

# Restaurant timelines

I attempt to get the restaurant timelines accross time.

```{r}
# df %>% 
#   filter(inspection_year > 2014) %>% 
#   filter(gradable == "yes") %>% 
#   ggplot(aes(x = inspection_date, y = score, group = restaurant_id)) +
#   geom_line(alpha = 0.3) +
#   facet_wrap(~borough)
```

Ok, it's very difficult to read And long to plot. Let's concentrate on 2017 and 2018. And gradable visits.

```{r}
df %>% 
  filter(inspection_year > 2016) %>% 
  filter(inspection_year < 2019) %>% 
  filter(gradable == "yes") %>% 
  ggplot(aes(x = inspection_date, y = score, group = restaurant_id)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~borough)
```


It's still horrible. Let's have a look at what it's happening in Manhattan.

```{r}
df %>% 
  filter(borough =="MANHATTAN") %>% 
  filter(inspection_year > 2016) %>% 
  filter(gradable == "yes") %>% 
  ggplot(aes(x = inspection_date, y = score, 
             group = restaurant_id)) +
  geom_line(alpha = 0.3)
```


Ok, obviously I won't go far with all restaurants. Let's go for proportion of A, B and C at a given time.

```{r}
df %>% 
  drop_na(grade) %>% 
  filter(grade %in% c("A", "B", "C")) %>% 
  filter(inspection_year > 2014 & inspection_year < 2019) %>% 
  group_by(inspection_floored) %>% 
  add_tally() %>% 
  add_count(inspection_floored, grade) %>% 
  distinct(inspection_floored, grade, n, nn) %>%
  mutate(prop = nn / n) %>% 
  ggplot(aes(x = inspection_floored, y = prop, colour = grade)) +
  geom_line(size = 2) +
  labs(title = "Proportion of each restaurant grade accross time (2015 to 2018)",
       subtitle = "A is good, B and C are not!",
       x = "Time",
       colour = "Inspection grade") +
  theme_light() +
  scale_colour_viridis(option = "viridis", discrete = TRUE)
```

Ok, so I do not see the trend of the original 538 article. Like them, I removed the other grades. I allowed a longer time period, but I don't really see the same signla in the period they looked at. Maybe they kept the restaurants with no current grades in the lot?