---
title: "Untitled"
author: "Mathilde Mousset"
date: "26 janvier 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(viridis)
library(hrbrthemes)

library(sf)
library(tigris)
library(gganimate)

load("././data.RData")
```
and I want to plot things on maps! I used R to make map of my populations, or field work samples, but this dataset seems like a good occasion to learn more on visualizing data on maps. So my aims for this post are: pretty graphs and pretty maps.





# Exploration

I am interested in pre-trial data, so the number or percentage of people who go through jail before their trial. This time I will investigate the `pretrial_population` dataset. This dataset has nine columns:  
- `year`: the year of the census, from 1970  
- `state`: abbreviated names of states  
- `county_name`: name of the county within state    
- `urbanicity`: as explained in the part 1, urbanicity represent the size and type of county (small and rural vs bigger and urban to very big urban areas)    
- `region`: the region delimitation in the USA    
- `dividion`: same as before, but finer  
- `pop_category`: labels that describe some aspects of the population  
- `population`: number of persons in the county 
- `pretrial_population`: number of persons in jail

So the first thing will be to get the ratio of people in jail compared to the county population.

```{r prison_rate}
pretrial_population <- pretrial_population %>% 
  mutate(incarceration_rate = pretrial_population*1000/population)
```

I also built a state summary of the jail population

```{r by_state}
pretrial_population_state <- pretrial_population %>%
  group_by(year, state, pop_category) %>% 
  summarise(population = sum(population, na.rm = TRUE),
            pretrial_population = sum(pretrial_population, na.rm = TRUE)) %>% 
  mutate(incarceration_rate = (pretrial_population / population) * 1000) %>% 
  ungroup()
```



Now, I will explore the missing data. I am afraid I was spoiled by other's explorations of this dataset, I already know that there are many missing data. Let's get a feel about them.

```{r}
pretrial_population
```



# Get some maps going



## Per state, Total population

From / Inspired from: http://www.rjsdatascience.com/post/prison-rate/


```{r}
library(tigris)
state_sf <- states(class = "sf")

state_prison_map <- state_sf %>% 
  left_join(pretrial_population_state, by = c("STUSPS" = "state"))

# Removing Hawai and Alaska to make mapping easier
state_prison_map %>% 
  filter(year == 2015, 
         pop_category == "Total", 
         !STUSPS %in% c("AK", "HI")) %>% 
  ggplot() +
  geom_sf(aes(fill = incarceration_rate)) +
  labs(title = "Incarceration rates in jail in 2015",
       subtitle = "Data: USA, 2015, rates as 1 per 1000 individuals",
       fill = "Incarceration rates\n(1 per 1000)") +
  theme_ipsum_rc() +
  scale_fill_viridis_c(direction = -1)
```

Animation in time


```{r}
animated_jail_states <- state_prison_map %>% 
  filter(year %in% c(2014:2016)) %>% 
  filter(pop_category == "Total", 
         !STUSPS %in% c("AK", "HI")) %>% 
  ggplot() +
  geom_sf(aes(fill = incarceration_rate)) +
  transition_time(as.integer(year)) +
  labs(title = "Incarceration rates in jail in {frame_time}",
       subtitle = "Data: USA, {frame_time}",
       fill = "Rate in jail\n(1 per 1000)",
       x = "",
       y = "") +
  theme_ipsum_rc() +
  scale_fill_viridis_c()

animate(animated_jail_states)
anim_save("././map_ani.gif")
```






# Other maps

```{r}
library(ggplot2)
library(usmap)

map_data("usa", region = "state") %>% 
  tbl_df()

state.fips
state.abb
state.area
state.division
state.name

map_data("state")
```

