---
title: "Untitled"
author: "Mathilde Mousset"
date: "19 février 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# From Otho

Also, I will learn from Otho Mantegazza, who did some nice clustering. Since I very rarelly used these tools, I thought it would be good to try to replicate his code.

https://otho.netlify.com/post/government-funding/

Otho looked at correlations, and then clustered agencies by similarity of founding patterns


## Correlations

```{r}
rd_wide <- 
  feder_spendings %>% 
  select(department, year, rd_budget) %>% 
  tidyr::spread(key = department,
                value = rd_budget)
```


```{r}
# correlation of everything against everything
rd_wide %>%
  column_to_rownames("year") %>%
  # Test correlation
  cor() %>% 
  # visualize
  as_tibble(rownames = "agency") %>% 
  gather(DHS:VA, key = "agency_2", value = "corr") %>% 
  ggplot(aes(x = agency,
             y = agency_2,
             fill = corr)) +
  geom_tile() + 
  scale_fill_viridis_c() +
  theme(aspect.ratio = 1,
        axis.text.x = element_text(angle = 90, hjust = .5))
```

## Clustering

>> With clusters, we want to capture the behaviour of our variables. Meaning that we want to capture how one variable behaves compared to another: when one variable increases in value, does the other do the same?  

>> So, to make variables comparable, we must scale them.

>> Probably, the most common scaling is by z-scores. But, because z-scores are centered at 0. This introduces negative numbers, but spendings for R&D agencies are (hopefully ;) ) never negative.

>> Thus, I decided to rescale spendings for each agency between 0 and 1, in this way we can make them comparable and present them in a more intuitive way. Moreover, this kind of scaling preserves 0 as anchored point.

```{r}
# scale each variable between 0 and 1
rd_wide_01 <- 
  rd_wide %>%
  mutate_at(vars(-contains("year")),
            ~scales::rescale(., to = c(0,1),
                             from = c(0, max(.))))


# this is how it looks after scaling
rd_wide_01 %>% print()
```

>> After scaling, I tried hierachical clustering. But, clustering functions such as hclust() or kmeans(), by default, cluster the rows of a dataset: we must to modify the format of the data again.

>> This time we transpose the data with t(), to exchange rows and columns.

```{r}
# For clustering
# we need to transpose the data
# because dist() %>% clusters the columns
rd_for_clust <-   
  rd_wide_01 %>%
  column_to_rownames("year") %>%
  t() 

rd_hclust <-
  rd_for_clust %>% 
  dist() %>%
  hclust() 
```

Let's visualize the results

```{r}
rd_hclust %>% plot()
```

```{r}
rd_for_clust %>% 
  superheat::superheat(row.dendrogram = TRUE,
                       left.label.text.size = 3,
                       bottom.label.text.angle = 90,
                       bottom.label.text.size = 3,
                       grid.hline.col = "grey40",
                       grid.vline.col = "grey40")
```

